Designing an effective intervention of motivational interview using sequence classification methods

### 2017-02-27 paper/AMIA/main.tex
update markov chain subsection

Our previous studies \cite{kotov2015interpretable, hasan2016study} explored several machine learning methods for the automatic annotation of clinical interview fragments, which uses a codebook \cite{carcone2013provider} with 115 different behavior codes. In this study, probabilistic models are employed for the analysis of sequential data to determine provider-patient communication sequences that are likely to translate into change talk and commitment language. To the best of our knowledge, there have been very limited studies on sequence classification in motivational interview such as patient-provider dialogue. A sequential analysis \cite{eide2004physician} was performed on physician-patient dialogue to find the relationship between physician and patient behaviors associated with patients' exibiting cues. Currently, a very similar work \cite{jaber2016multi} was done on education data, where a sequence classification model was developed to predict the role of a student in a project team based on their communication patterns. Although Hidden Markov Model was originally applied to speech recognition \cite{rabiner1989tutorial}, this is one of the most successful method \cite{mutsam2016maximum, eickeler1998hidden, won2004training, chai2001folk} for sequence classification. Therefore, we utilize Markov Chain and Hidden Markov Model for the classification of utterance sequences.

@@

\subsection*{\textit{System overview}}
The pipeline consists of two stages: training and testing. Prior to the training stage, we preprocess the collected clinical interview transcripts by performing stemming, punctuation removal, word segmentation and tokenization. Features are then extracted from the preprocessed data. During this stage, previous label and LIWC features are used in conjunction with the lexical features to create the feature vectors. After that, classifiers are trained on the feature vectors extracted from the training samples and their associated annotations. In the testing stage, after creation of feature vectors, the previously trained classifiers predict the label of each utterance in the testing sample. Finally, performance of different classifiers is evaluated by calculating standard metrics such as precision, recall, F-score (F1), kappa measure and accuracy. Specifically, we evaluated the performance of the following state-of-the-art supervised machine learning methods.

\textbf {Markov Model}: In probability theory, a markov model \cite{} is a stochastic model used to model randomly changing systems where it is assumed that future states depend only on the present state and not on the sequence of events that preceded it (that is, it assumes the markov property). Generally, this assumption enables reasoning and computation with the model that would otherwise be intractable. For the sequential analysis, we built two markov models $M$ and $\overline{M}$ describing provider strategies and patient responses in case of successful ($M$) and unsuccessful ($\overline{M}$) motivational interviews. A markov model $M$ can be represented as a weighted directed graph $G = (V, E, p)$, in which:

@@

@@

Direction of future work include the exploration of other methods such as BLAST \cite{altschul1990basic} and Strand \cite{drew2014strand} for the sequence classification.

>>>>

Our previous studies \cite{kotov2015interpretable, hasan2016study} explored several machine learning methods for the automatic annotation of clinical interview fragments, which uses a codebook \cite{carcone2013provider} with 115 different behavior codes. In this study, probabilistic models are employed for the analysis of sequential data to determine provider-patient communication sequences that are likely to translate into change talk and commitment language. To the best of our knowledge, there have been very limited studies on sequence classification in motivational interview such as patient-provider dialogue. A sequential analysis \cite{eide2004physician} was performed on physician-patient dialogue to find the relationship between physician and patient behaviors associated with patients' exibiting cues. Currently, a very similar work \cite{jaber2016multi} was done on education data, where a sequence classification model was developed to predict the role of a student in a project team based on their communication patterns. Although Hidden Markov Model was originally applied to speech recognition \cite{rabiner1989tutorial}, this is one of the most successful method \cite{mutsam2016maximum, eickeler1998hidden, srivastava2007hmm, won2004training, chai2001folk} for sequence classification. In addition, a k-order Markov Model is applied on genomic research \cite{yakhnenko2005discriminatively} to classify protein and text sequence data.  Therefore, we utilize Markov Chain and Hidden Markov Model for the classification of utterance sequences.

@@

\subsection*{\textit{Sequence classification techniques}}
Generally, a sequence is an ordered list of events. In our project, event is a behavior code represented as a symbolic value such as 117 (low uptake, positive), 331 (open-ended question), etc.  Given a sequence of behavior codes represented as $S_i = \{x_1, x_2,...,x_n\}$ and a set of class labels $L = \{l_1, l_2,...,l_m\}$, the task of sequence classification is to learn a sequence classifier C, which is a function mapping a sequence $S_i$ to a class label $l_i \in L$, written as, $C : S_i \to l_i, l_i \in L$. Here, L is fixed and known in advances. In our experiment, known labels are ``successful'' and ``unsuccessful'' motivational interview.

\textbf {Markov Chain}: In probability theory, a markov model \cite{} is a stochastic model used to model randomly changing systems where it is assumed that future states depend only on the present state and not on the sequence of events that preceded it (that is, it assumes the markov property). Generally, this assumption enables reasoning and computation with the model that would otherwise be intractable. For the sequential analysis, we built two markov models $M$ and $\overline{M}$ describing provider strategies and patient responses in case of successful ($M$) and unsuccessful ($\overline{M}$) motivational interviews. A markov model $M$ can be represented as a weighted directed graph $G = (V, E, p)$, in which:

@@

The Markov Model we have discussed so far is referred to as the first order Markov Model, since we only look one code in the past to compute the transition probability. A generalization of the first order markov model is the $k^{th}$ order markov chain in which the transition probability for a particular symbol $x_i$ is computed by looking at the k preceding symbols. Thus, $k^{th}$ order markov chain will have $N^{k}$ states each associated with a sequence of k symbols. In our experiment, we also used second order markov model. Hence, our model have $N^2$ states with transitional probability matrix of size $N^2 \times N$.

\textbf {Hidden Markov Model}: Hidden Markov Models are widely used for sequence analysis because of their ability to incorporate dependencies among elements in sequence.

@@

Direction of future work include the exploration of other methods such as BLAST \cite{altschul1990basic} and Strand \cite{drew2014strand} for the classification of utterance sequences.
